---
title: "JSC370 Final Project Plots"
author: "Richard Guo"
output:
  html_document: default
always_allow_html: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = TRUE, echo  = FALSE, warning = FALSE, include = TRUE)
suppressPackageStartupMessages({
  library(ggplot2)
  library(plotly)
  library(dplyr)
  library(nlme)
  library(mgcv)         
  library(randomForest)
  library(xgboost)
  library(httr)
  library(lubridate)
  library(htmlwidgets)
})
```
[Download the full report (PDF)](JSCFinal.pdf)
For better reading experience, [Download the full report (HTML)](JSCFinal.html)
```{r, echo = FALSE}
source("finalprojscript.R")
library(plotly)
library(dplyr)

final_table <- final_table %>%
  mutate(month = factor(month, levels = month.name))  # Ensure months are in order


  # Create interactive plot with dual y-axes
p1 <-  plot_ly(final_table) %>%
    add_lines(
      x = ~month, y = ~num_crimes,
      name = "Number of Crimes",
      yaxis = "y1",
      line = list(color = "blue")
    ) %>%
    add_lines(
      x = ~month, y = ~avg_temp,
      name = "Average Temperature (°C)",
      yaxis = "y2",
      line = list(color = "red")
    ) %>%
    layout(
      title = "Monthly Average Temperature and Number of Crimes",
      xaxis = list(title = "Month"),
      yaxis = list(title = "Number of Crimes"),
      yaxis2 = list(
        title = "Average Temperature (°C)",
        overlaying = "y",
        side = "right"
      ),
      legend = list(x = 0.1, y = 1)
    )

# === Variable Importance Plots ===
# RF

var_labels <- c(
  temperature = "Temperature",
  humidity = "Humidity",
  precipitation = "Precipitation",
  snowfall = "Snowfall",
  windsp = "Wind Speed",
  visibility = "Visibility",
  dew_point = "Dew Point"
)


library(plotly)

# Get importance and relabel
rf_imp <- importance(rf_model)
rf_df <- data.frame(
  Variable = rownames(rf_imp),
  Importance = rf_imp[, "%IncMSE"],
  stringsAsFactors = FALSE
)
rf_df$Variable <- var_labels[rf_df$Variable]  # Apply readable labels

# Plot
p_rf <- ggplot(rf_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest Variable Importance",
       x = "", y = "% Increase in MSE", caption = "Fig 6: Variable Importance for Random Forest") +
  theme_minimal()

p2 <-  ggplotly(p_rf)
# XGB
imp_xgb <- xgb.importance(model = xgb_model)

# Relabel feature names
imp_xgb$Feature <- var_labels[imp_xgb$Feature]

# Plot
p_xgb <- ggplot(imp_xgb, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(title = "XGBoost Variable Importance", x = "", y = "Gain", caption = "Fig 7: XGBoost Variable Importance") +
  theme_minimal()


p3 <-  ggplotly(p_xgb)

```


```{r, echo=FALSE}
p1
```
Website Fig 1: Interactive plot of average temperature and number of crimes per month (Toronto, 2024). The relationship is magnified when the y-axis for number of crimes is scaled to fit the range of values. Both trends peak during the summer months, indicating an increase in crimes during warmer weather.
```{r, echo=FALSE}
p2
```
Website Fig 2: Variable importance for the Random Forest Model. The most important variables are humidity, visibility, and temperature. Note that this model only included weather-related factors, and other economic, social, or demographic factors may (and likely have) a bigger impact on crime rates overall.
```{r, echo=FALSE}
p3
```
Website Fig 3: Variable importance for the XGBoost Model. The most important variables are wind speed, temperature, and visibility. Wind speed is much more important in this model than the random forest model, while temperature and visibility are present in both models. Both models agree on snowfall being the least important variable, followed by precipitation.
