---
title: "Lab 11 - Interactive Visualization"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = TRUE, include  = TRUE)
```

# Learning Goals

- Read in and process Starbucks data.
- Create interactive visualizations of different types using `plot_ly()` and `ggplotly()`.
- Customize the hoverinfo and other plot features.
- Create a Choropleth map using `plot_geo()`.

# Lab Description

We will work with two Starbucks datasets, one on the store locations (global) and one for the nutritional data for their food and drink items. We will do some text analysis of the menu items.

# Deliverables

Upload an html file to Quercus and make sure the figures remain interactive.

# Steps

### 0. Install and load libraries

```{r message=FALSE, echo=FALSE, warning=FALSE}

library(tidyverse)
library(plotly)
library(widgetframe)
library(tidytext)


```

### 1. Read in the data

- There are 4 datasets to read in, Starbucks locations, Starbucks nutrition, US population by state, and US state abbreviations. All of them are on the course GitHub.

```{r, echo = TRUE}

sb_locs <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/starbucks-locations.csv")

sb_nutr <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/starbucks-menu-nutrition.csv")

usa_pop <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/us_state_pop.csv")

usa_states<-read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/states.csv")
```

### 2. Look at the data

- Inspect each dataset to look at variable names and ensure it was imported correctly.


```{r, echo = TRUE}
sb_locs
sb_nutr
usa_pop
usa_states
```

### 3. Format and merge the data

- Subset Starbucks data to the US.
- Create counts of Starbucks stores by state.
- Merge population in with the store count by state.
- Inspect the range values for each variable. 

```{r, echo=TRUE}

# Subset Starbucks locations to only those in the US
sb_usa <- sb_locs |> filter(Country == "US")

# Count Starbucks stores by state
sb_locs_state <- sb_usa |>
  group_by(`State/Province`) |>
  summarize(store_count = n())

# Rename the state abbreviation column in the states dataset
usa_states <- usa_states |> rename(state = State, abbreviation = Abbreviation)

# Merge state population data with state abbreviations
usa_pop_abbr <- full_join(usa_pop, usa_states, by = c("state" = "state"))

# Merge the Starbucks store count data with the population dataset
sb_locs_state <- full_join(sb_locs_state, usa_pop_abbr, 
                           by = c("State/Province" = "abbreviation"))

# Inspect the range of values
summary(sb_locs_state)

  


```

### 4. Use `ggplotly` for EDA

Answer the following questions:

- Are the number of Starbucks proportional to the population of a state? (scatterplot)

- Is the caloric distribution of Starbucks menu items different for drinks and food? (histogram)

- What are the top 20 words in Starbucks menu items? (bar plot)

```{r, include = TRUE}
# Scatterplot of Starbucks stores vs. population
p1 <- sb_locs_state |> 
  ggplot(aes(x = population, y = store_count, text = state)) + 
  geom_point(color = "darkgreen", alpha = 0.6) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Starbucks Stores vs. State Population",
       x = "Population",
       y = "Number of Starbucks Stores") +
  theme_minimal()

ggplotly(p1, tooltip = "text")



```

- 4a) Answer: There is a strong positive correlation between population and number of starbucks stores.

```{r, echo = TRUE}

# Ensure there's a category column to distinguish drinks vs. food
p2 <- sb_nutr |> 
  ggplot(aes(x = Calories, fill = Category)) + 
  geom_histogram(bins = 30, position = "identity", alpha = 0.6) + 
  labs(title = "Caloric Distribution: Drinks vs. Food",
       x = "Calories",
       y = "Count",
       fill = "Category") +
  theme_minimal()

ggplotly(p2)


```

- 4b) Answer: Food, on average, has higher calories. There is a large amount of drinks with 0 calories.

```{r, echo = TRUE}

# Tokenizing menu items into words
top_words <- sb_nutr |> 
  unnest_tokens(word, `Item`) |>  # Assuming 'Beverage_category' contains item names
  anti_join(stop_words, by = "word") |>  # Remove common stopwords
  count(word, sort = TRUE) |> 
  slice_max(n, n = 20)

# Bar plot of top 20 words
p3 <- top_words |> 
  ggplot(aes(x = reorder(word, n), y = n, fill = word)) + 
  geom_col(show.legend = FALSE) + 
  coord_flip() +
  labs(title = "Top 20 Words in Starbucks Menu Items",
       x = "Word",
       y = "Count") +
  theme_minimal()

ggplotly(p3)

```

- 4c) Answer: Top 20 words are iced, tazo, bottled, sandwich, chocolate, coffee, tea, starbucks, egg, protein, cake, black, white, mocha, latte, cookie, chicken, vanilla, salad, shaken, organic, and macchiato.

### 5. Scatterplots using `plot_ly()`

- Create a scatterplot using `plot_ly()` representing the relationship between calories and carbs. Color the points by category (food or beverage). Is there a relationship, and do food or beverages tend to have more calories?

```{r, echo = TRUE}
# Scatterplot of Calories vs. Carbs using plot_ly
plot_ly(data = sb_nutr, 
        x = ~`Carb. (g)`, 
        y = ~Calories, 
        color = ~Category,  # Coloring by food or beverage category
        type = "scatter", 
        mode = "markers",
        marker = list(size = 8, opacity = 0.6)) %>%
  layout(title = "Calories vs. Carbohydrates in Starbucks Menu Items",
         xaxis = list(title = "Carbohydrates (g)"),
         yaxis = list(title = "Calories"),
         legend = list(title = list(text = "Category")))

```


- 5a) Answer: There seems to be a positive correlation betweenc carbs and calories. Food, on average, has higher calories than drinks.

- Repeat this scatterplot but for the items that include the top 10 words. Color again by category, and add hoverinfo specifying the word in the item name. Add layout information to title the chart and the axes, and enable `hovermode = "compare"`.
- What are the top 10 words and is the plot much different than above?
  
```{r, echo = TRUE}
# Tokenizing menu items into words and counting occurrences
top_10_words <- sb_nutr |> 
  unnest_tokens(word, `Item`) |>  # Assuming 'Beverage_category' contains item names
  anti_join(stop_words, by = "word") |>  # Remove common stopwords
  count(word, sort = TRUE) |> 
  slice_max(n, n = 10) |> 
  pull(word)  # Extract as a vector

# Create a regex pattern to match any of the top 10 words
pattern <- paste0("\\b(", paste(top_10_words, collapse = "|"), ")\\b")

# Filter for menu items that contain at least one of the top 10 words
sb_nutr_filtered <- sb_nutr |> 
  filter(str_detect(`Item`, regex(pattern, ignore_case = TRUE))) |> 
  mutate(matching_word = str_extract(`Item`, regex(pattern, ignore_case = TRUE)))


plot_ly(data = sb_nutr_filtered, 
        x = ~`Carb. (g)`, 
        y = ~Calories, 
        color = ~Category,  # Coloring by food or beverage category
        text = ~paste("Word:", matching_word, "<br>Item:", `Item`), 
        type = "scatter", 
        mode = "markers",
        marker = list(size = 8, opacity = 0.7)) %>%
  layout(title = "Calories vs. Carbohydrates for Items Containing Top 10 Words",
         xaxis = list(title = "Carbohydrates (g)"),
         yaxis = list(title = "Calories"),
         legend = list(title = list(text = "Category")),
         hovermode = "compare")  # Enables comparing hover tooltips

```

- 5b) Answer: The same trend is present.



### 6. `plot_ly` Boxplots

- Create a boxplot of all of the nutritional variables in groups by the 10 item words.
- Which top word has the most calories? Which top word has the most protein?

```{r, echo = TRUE}
# Convert data to long format for boxplot visualization
sb_nutr_long <- sb_nutr_filtered |> 
  select(matching_word, Calories, `Fat (g)`, `Carb. (g)`, `Fiber (g)`, `Protein (g)`) |>  # Select only nutrition columns
  pivot_longer(cols = Calories:`Protein (g)`,  # Convert all nutrient columns into long format
               names_to = "Nutrient", 
               values_to = "Value")
plot_ly() |> 
  # Boxplot for Calories (Primary Y-Axis)
  add_trace(data = sb_nutr_long |> filter(Nutrient == "Calories"), 
            x = ~matching_word, 
            y = ~Value, 
            type = "box", 
            name = "Calories",
            marker = list(color = "red"),
            yaxis = "y1") |> 
  
  # Boxplots for Other Nutrients (Secondary Y-Axis)
  add_trace(data = sb_nutr_long |> filter(Nutrient != "Calories"), 
            x = ~matching_word, 
            y = ~Value, 
            type = "box", 
            name = ~Nutrient,
            color = ~Nutrient,
            yaxis = "y2") |> 

  layout(title = "Nutritional Distribution of Top 10 Starbucks Menu Words",
         xaxis = list(title = "Top 10 Words in Menu Items"),
         
         # Primary Y-Axis (Calories)
         yaxis = list(title = "Calories", 
                      side = "left",
                      showgrid = FALSE),
         
         # Secondary Y-Axis (Other Nutrients)
         yaxis2 = list(title = "Other Nutrients (g)", 
                       overlaying = "y", 
                       side = "right",
                       showgrid = FALSE),
         
         boxmode = "group")

```

- 6) Answer: Sandwich has the most calories and the most protein.

### 7. 3D Scatterplot

- Create a 3D scatterplot between Calories, Carbs, and Protein for the items containing the top 10 words
- Do you see any patterns (clusters or trends)?

```{r, echo = TRUE}
# Create a 3D scatterplot
plot_ly(data = sb_nutr_filtered, 
        x = ~`Carb. (g)`, 
        y = ~Calories, 
        z = ~`Protein (g)`, 
        color = ~matching_word,  # Color by top word
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 5, opacity = 0.6)) %>%
  layout(title = "3D Scatterplot of Calories, Carbs, and Protein for Top 10 Words",
         scene = list(
           xaxis = list(title = "Carbohydrates (g)"),
           yaxis = list(title = "Calories"),
           zaxis = list(title = "Protein (g)")
         ))


```

- 7) Answer: There is a clear positive correlation between calories, protein, and carbs, and a small cluster of items with 0 proteins having a positive correlation ebtween calories and carbs.


### 8. `plot_ly` Map

- Create a map to visualize the number of stores per state, and another for the population by state. Add custom hover text. Use subplot to put the maps side by side.
- Describe the differences if any.

```{r, echo = TRUE}



# Create hover text
# Create hover text
sb_locs_state$hover <- with(sb_locs_state, paste("Number of Starbucks: ", store_count, '<br>', "State: ", state, '<br>', "Population: ", population))


# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('steelblue')
)

# Make sure both maps are on the same color scale
shadeLimit <- 125


# Create the map

# Create a map for the number of stores
map1 <- plot_geo(sb_locs_state, locationmode = 'USA-states') %>%
  add_trace(
    z = ~store_count,
    locations = ~`State/Province`,
    color = ~store_count,
    colors = 'Reds',
    hoverinfo = 'text',
    text = ~hover
  ) %>%
  layout(
    title = "Number of Starbucks Stores per State",
    geo = set_map_details
  )
# Create a map for the population
map2 <- plot_geo(sb_locs_state, locationmode = 'USA-states') %>%
  add_trace(
    z = ~population,
    locations = ~`State/Province`,
    color = ~population,
    colors = 'Blues',
    hoverinfo = 'text',
    text = ~hover
  ) %>%
  layout(
    title = "Population per State",
    geo = set_map_details
  )
# Place maps side by side
subplot(map1, map2, nrows = 1, shareX = TRUE, shareY = TRUE) %>%
  layout(
    title = "Starbucks Stores and Population by State"
  )

```

- 8) Answer: The maps seem to follow the same pattern, with California having a large population as well as a large number of stores.
